Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-07-09 text/manuscript-full-KS.qmd
revisions rq1 LA

Due to the large number of Geomarkets in urban Southern California, we distinguish Los Angeles -- 11 Geomarkets, results shown here -- from Orange County (four Geomarkets) and San Diego (three Geomarkets), results for which are shown in the [Online Appendix](https://ozanj.github.io/cb_geo/#/results). The 11 Geomarkets within the Los Angeles metropolitan area are illustrated in @fig-la-maps by total population across the three census periods. The Los Angeles metropolitan area is one of the only multi-nuclei models in the country that is formed when urban growth occurs around multiple centers that serve different functions rather than expanding from a single central business district core (CITE- Harris and Ullman 1945). As depicted in @fig-la-maps, Geomarkets closely follow this overall metropolitan model. Hollywood and Wilshire, South and South Central Los Angeles, and East Los Angeles Geomarkets are centrally located in the metropolitan area followed by an outer core ring made up by the Glendale and Pasadena, San Fernando Valley-East, West Los Angeles and West Beach, South Bay, and Long Beach Geomarkets. The San Fernando Valley-West Geomarket is located in the northwest part of the region, whereas Covina and West Covina as well as the Riverside, San Bernadino, and Ontario Geomarkets are located inland, comprising the Inland Empire.

The population growth in the Los Angeles metropolitan area increased from approximately 8 Million in 1980 to more than 12 Million in 2020. @fig-la-maps showcases population growth by Geomarkets over the three census periods reflect the urban sprawl forming the multi-nuclei model. The most pronounced increases occurred in South and South Central Los Angeles Geomarket, as well as the outer/inland Geomarkets of San Fernando Valley-West, Riverside and San Bernadino, and Ontario. The South and South Central Los Angeles Geomarket grew from approximately 1.2 to 1.6 Million from 1980 to 2020. In contrast, the San Fernando Valley-West and Riverside and San Bernadino, and Ontario Geomarkets, respectively, grew from approximately 700,000 to 1.2 Million and from 1.1 Million to nearly 3 Million during this time.

@fig-la-rq1-race presents the racial and ethnic composition of each Geomarket within the Los Angeles metropolitan area from 1980 to 2020. One data caveat here is that the Asian race/ethnicity category was not available in 1980. However, over the three decennial census periods, the proportion of White residents declined across all Geomarkets, while Asian and Hispanic populations generally increased. In 1980, all Geomarkets - except for East Los Angeles, South and South Central Los Angeles, and Hollywood and Wilshire- were predominantly White, with percentages ranging from 62% White in Long Beach to 78% in the San Fernando Valleys (West and East) Geomarkets. By 2020, however, only the West Los Angeles and West Beach Geomarket remained predominantly White (62%). Remanining Geomarkets exhibited a more racially/ethnic diverse population by 2020, including San Fernando Valley-West- (44% White, 4% Black, 39% Hispanic; 11% Asian Pacific Islander), San Fernando Valley -East (43% White, 4% Black, 41% Hispanic; 8% Asian Pacific Islander), Glendale and Pasadena (35% White, 3% Black, 37% Hispanic; 21% Asian Pacific Islander), South Bay (29% White, 12% Black, 33% Hispanic; 21% Asian Pacific Islander), Long Beach (26% White, 11% Black, 46% Hispanic; 13% Asian Pacific Islander), Covina and West Covina (19% White, 3% Black, 52% Hispanic; 24% Asian Pacific Islander), and Riverside, San Bernardino, and Ontario (28% White, 7% Black, 54% Hispanic; 6% Asian Pacific Islander). However, the South and South Central Los Angeles and Hollywood and Wilshire Geomarkets experienced substantial declines in the share of Black residents from 1980 to 2020, dropping from 44% to 18% and 18% to 9%, respectively.

@@

@fig-la-rq1-ses presents trends in income, poverty, and educational attainment in the Los Angeles metropolitan area from 1980 to 2020, demonstrating the South and South Central Los Angeles, East Los Angeles, and Hollywood and Wilshire Geomarkets remained at the lowest socioeconomic levels across these indicators. For instance, the median household income in the South and South Central Los Angeles Geomarket increased modestly from \$57,000 in 1980 to \$59,000 in 2000, reaching \$67,000 by 2020. In comparison, East Los Angeles and Hollywood and Wilshire recorded slightly higher median incomes in 2020 at \$76,000 and \$84,000, respectively. These figures remain substantially lower than that of the most affluent Geomarket, West Los Angeles and West Beach, where the median household income reached \$137,000 in 2020.

@@

```{r fig-la-maps, echo = FALSE, fig.align = 'center', fig.cap = "Los Angeles Worth Geomarkets, Total Population 1980-2020", out.width = "100%"}
knitr::include_graphics('./../assets/images/la-maps.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-la-rq1-race, echo = FALSE, fig.align = 'center', fig.cap = "Racial/Ethnic Composition of Los Angeles Area Geomarkets, 1980-2020", out.width = "100%"}
knitr::include_graphics('./../assets/images/rq1_la_race.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-la-rq1-ses, echo = FALSE, fig.align = 'center', fig.cap = "Socioeconomic Characteristics of Los Angeles Area Geomarkets, 1980-2020", out.width = "100%"}
knitr::include_graphics('./../assets/images/rq1_la_ses.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

>>>>

Due to the large number of Geomarkets in urban Southern California, we distinguish Los Angeles -- 11 Geomarkets, results shown here -- from Orange County (four Geomarkets) and San Diego (three Geomarkets), results for which are shown in the [Online Appendix](https://ozanj.github.io/cb_geo/#/results). The 11 Geomarkets within the Los Angeles metropolitan area are illustrated in @fig-la-maps by total population across the three census periods. The Los Angeles metropolitan area is one of the only multi-nuclei models in the country that is formed when urban growth occurs around multiple centers that serve different functions rather than expanding from a single central business district core [@Harris1945]. As depicted in @fig-la-maps, Geomarkets closely follow this overall metropolitan model. Hollywood and Wilshire, South and South Central Los Angeles, and East Los Angeles Geomarkets are centrally located in the metropolitan area followed by an outer core ring made up by the Glendale and Pasadena, San Fernando Valley-East, West Los Angeles and West Beach, South Bay, and Long Beach Geomarkets. The San Fernando Valley-West Geomarket is located in the northwest part of the region, whereas Covina and West Covina as well as the Riverside, San Bernadino, and Ontario Geomarkets are located inland, comprising the Inland Empire.

The population growth in the Los Angeles metropolitan area increased from approximately 8 Million in 1980 to more than 12 Million in 2020. @fig-la-maps showcases population growth by Geomarkets over the three census periods reflect the urban sprawl forming the multi-nuclei model. The most pronounced increases occurred in South and South Central Los Angeles Geomarket, as well as the outer/inland Geomarkets of San Fernando Valley-West, Riverside and San Bernadino, and Ontario. The South and South Central Los Angeles Geomarket grew from approximately 1.1 to 1.6 Million from 1980 to 2020. In contrast, the San Fernando Valley-West and Riverside and San Bernadino, and Ontario Geomarkets, respectively, grew from approximately 580,000 to nearly 1 Million and from 1.1 Million to 3.3 Million during this time.

@fig-la-rq1-race presents the racial and ethnic composition of each Geomarket within the Los Angeles metropolitan area from 1980 to 2020. One data caveat here is that the Asian race/ethnicity category was not available in 1980. However, over the three decennial census periods, the proportion of White residents declined across all Geomarkets, while Asian and Hispanic populations generally increased. In 1980, all Geomarkets - except for East Los Angeles, South and South Central Los Angeles, and Hollywood and Wilshire- were predominantly White, with percentages ranging from 63% White in Covina and West Covina to 86% in the San Fernando Valley-East Geomarket. By 2020, however, only the West Los Angeles and West Beach (51%) and San Fernando Valley-East (59%) Geomarkets remained predominantly White. Remanining Geomarkets exhibited a more racially/ethnic diverse population by 2020, including San Fernando Valley-West- (40% White, 4% Black, 40% Hispanic; 13% Asian Pacific Islander), Glendale and Pasadena (35% White, 4% Black, 41% Hispanic; 17% Asian Pacific Islander), South Bay (32% White, 10% Black, 32% Hispanic; 22% Asian Pacific Islander), Long Beach (27% White, 10% Black, 47% Hispanic; 13% Asian Pacific Islander), Covina and West Covina (19% White, 4% Black, 57% Hispanic; 19% Asian Pacific Islander), and Riverside, San Bernardino, and Ontario (28% White, 7% Black, 53% Hispanic; 8% Asian Pacific Islander). However, the South and South Central Los Angeles and Hollywood and Wilshire Geomarkets experienced substantial declines in the share of Black residents from 1980 to 2020, dropping from 44% to 16% and 22% to 11%, respectively.

@@

@fig-la-rq1-ses presents trends in income, poverty, and educational attainment in the Los Angeles metropolitan area from 1980 to 2020, demonstrating the South and South Central Los Angeles, East Los Angeles, and Hollywood and Wilshire Geomarkets remained at the lowest socioeconomic levels across these indicators. For instance, the median household income in the South and South Central Los Angeles Geomarket increased modestly from \$51,000 in 1980 to \$57,000 in 2000, reaching \$61,000 by 2020. In comparison, East Los Angeles and Hollywood and Wilshire recorded slightly higher median incomes in 2020 at \$74,000 and \$70,000, respectively. These figures remain substantially lower than that of the most affluent Geomarket, West Los Angeles and West Beach, where the median household income reached \$117,000 in 2020.

@@

```{r fig-la-maps, echo = FALSE, fig.align = 'center', fig.cap = "Los Angeles Geomarkets, Total Population 1980-2020", out.width = "100%"}
knitr::include_graphics('./../assets/images/la-maps.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-la-rq1-race, echo = FALSE, fig.align = 'center', fig.cap = "Racial/Ethnic Composition of Los Angeles Area Geomarkets, 1980-2020", out.width = "100%"}

knitr::include_graphics('./../results/graphs/rq1/rq1_los_angeles_race.png')

#knitr::include_graphics('./../assets/images/rq1_la_race.png') [OLD RESULTS BASED ON ZIP GEOMARKETS]
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-la-rq1-ses, echo = FALSE, fig.align = 'center', fig.cap = "Socioeconomic Characteristics of Los Angeles Area Geomarkets, 1980-2020", out.width = "100%"}
knitr::include_graphics('./../results/graphs/rq1/rq1_los_angeles_ses.png')

#knitr::include_graphics('./../assets/images/rq1_la_ses.png') [OLD RESULTS BASED ON ZIP GEOMARKETS]
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```
